### 2022-02-14 paper.tex
in progress

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{amsmath,amsfonts,graphicx}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{multirow}
\usetikzlibrary {shapes.geometric}
\usetikzlibrary {positioning}

\begin{document}

\title{Monitoring Time Series With Missing Values:\\a Deep Bayesian Approach}

\author{Oshri Barazani\inst{1} \and David Tolpin\inst{2}}
\institute{PUB+ \and Ben-Gurion University of the Negev
\\ \email{\{oshribr,david.tolpin\}@gmail.com}}

\maketitle

\begin{abstract}
Systems are commonly monitored for health and security through collection and
streaming of multivariate time series. Advances in time series forecasting due
to adoption of multilayer recurrent neural network architectures make it
possible to forecast in high-dimensional time series, and identify and classify
novelties early, based on subtle changes in the trends. However, mainstream
approaches to multi-variate time series predictions do not handle well cases
when the ongoing forecast must include uncertainty, nor they are robust to
missing data. We introduce a new architecture for time series monitoring based
on combination of state-of-the-art methods of forecasting in high-dimensional
time series with full Bayesian handling of uncertainty. We demonstrate
advantage of the architecture for time series forecasting and novelty
detection, in particular with partially missing data, and empirically evaluate
and compare the architecture to state-of-the-art approaches on a real-world
data set.
\end{abstract}

\section{Introduction}

Evolving state of OT environments  is monitored through collection and
streaming of multivariate time series. Advances in time series forecasting due
to adoption of multilayer recurrent neural network architectures make it
possible to forecast in high-dimensional time series, and identify and
classify novelties (anomalies) early, based on subtle changes in the trends.
However, mainstream approaches to multi-variate time series predictions do not
handle well cases when the ongoing forecast must include uncertainty and shape
(possibly multimodal) of future distributions of observations. Stochastic
latent variable variants of high-dimensional time series models where
proposed, but so far have had to rely on sampling to account for uncertainty,
limiting the performance of data handling. In addition, main-stream methods of
multivariate time series forecasting are not well-suited for handling missing
data, while transient unavailability of some of the data sources (e.g. when a
sensor stops working, or a transport channel malfunctions) is  a common case
with monitoring of complex systems.  Combining state-of-the-art methods of
forecasting in high-dimensional time series (based on neural-network
architectures) with full Bayesian handling of uncertainty (e.g. with help of
deep probabilistic programming systems such as Pyro, Tensorflow Probability,
and others) will allow to achieve robust forecasting (a precondition for
anomaly detection) in high-dimensional time series with missing data.
Representing future states as inferred posterior probability distributions
will allow to identify and classify anomalies (e.g. hardware failures,
software errors, communication, malicious activity, cyberattacks, e.t.c.) on a
higher level of model abstraction — the latent variables of the model, and
achieve finer and more robust anomaly detection and characterization.

\section{Concept}

\begin{figure}
\begin{subfigure}{.64\textwidth}
\begin{tikzpicture}
\node [rectangle, draw] (rnn) {RNN};

\node (z_t) [above of=rnn] {$z_t$};
\node [trapezium, draw] (mlp) [above of=z_t] {MLP};
\draw [-] (rnn) to (z_t);
\draw [->] (z_t) to (mlp);

%output
\node (musigma) [above of=mlp] {$\mu_{t+1}, \sigma_{t+1}$};
\draw [->] (mlp) to (musigma);

%input
\node (x_t) [below of=rnn] {$x_t$};
\draw [->] (x_t) to (rnn);

\node (h_tm1) [left=12pt of rnn] {$h_{t-1}$};
\draw [->] (h_tm1) to (rnn);

\node (h_t) [right=12pt of rnn] {$h_{t}$};
\draw [-] (rnn) to (h_t);

\node (rnnnext) [rectangle, draw, right=12pt of h_t] {RNN};
\draw [->] (h_t) to (rnnnext);

\node (xnext) [below of=rnnnext] {$x_{t+1}$};
\draw [->] (xnext) to (rnnnext);

%sampling
\node (N) [gray,right=-4pt of xnext] {$\sim \mathcal{N}(\mu_{t+1}, \sigma_{t+1})$};
\draw [->, dashed,gray] (musigma) to [bend left] (N);

\end{tikzpicture}
\caption{Conventional model}
\label{fig:dt-conventional-model}
\end{subfigure}%
\begin{subfigure}{.36\textwidth}
\begin{tikzpicture}
\node [rectangle, draw] (rnn) {RNN};

\node (z_t) [above of=rnn] {$z_t$};
\node [trapezium, draw] (mlp) [above of=z_t] {MLP};
\draw [-] (rnn) to (z_t);
\draw [->] (z_t) to (mlp);

%output
\node (musigma) [above of=mlp] {$\mu_{t+1}, \sigma_{t+1}$};
\draw [->] (mlp) to (musigma);

%input
\node (x_t) [below of=rnn] {$\mu_t=x_t,\sigma_t=0$};
\draw [->] (x_t) to (rnn);

\node (h_tm1) [left=12pt of rnn] {$h_{t-1}$};
\draw [->] (h_tm1) to (rnn);

\node (h_t) [right=12pt of rnn] {$h_{t}$};
\draw [-] (rnn) to (h_t);

\node (rnnnext) [rectangle, draw, right=12pt of h_t] {RNN};
\draw [->] (h_t) to (rnnnext);

\node (xnext) [below of=rnnnext] {$\mu_{t+1}, \sigma_{t+1}$};
\draw [->] (xnext) to (rnnnext);

\node (F) [right=12pt of z_t] {};
\draw [-,dashed,gray] (musigma) to [bend left] (F);
\draw [->,dashed,gray] (F) to [bend right] (xnext);

\end{tikzpicture}
\caption{Model with uncertainty propagation}
\label{fig:dt-uncertainty-propagation}
\end{subfigure}
\caption{Time series models}
\label{fig:dt-time-series-models}
\end{figure}

There is a range of neural recurrent models of varying complexity to deal with
time series prediction. Most models include a recurrent unit which threads the
state through the time steps, accepts data as inputs and produces next step
predictions as outputs. The simplest model is an RNN with a fully-connected
readout layer to produce predictions (Figure~\ref{fig:dt-conventional-model}).
RNN can be based on GRU, LSTM, or another variant, and is often multi-layer.
Architectures  may also include intermediate modules and sampling-based
variational layers.

Variants are stochastic RNNs, variational RNNs, etc. The
overall architecture stays almost the same, with more
connections, intermediate modules and sampling-based variational
layers.

\paragraph{Input and Output} This architecture normally accepts data vectors
and outputs vectors of predictions of means and standard deviations (the output
is twice as wide as the input). Then, to generate a forecast, one samples from
$$\mathcal{N}(\mu, \sigma)$$ for each component, and feeds the sample as the
input to the next time step.

\paragraph{Training} The model is trained to maximize probability of
prediction. In the most basic case, a single step is predicted for each time
step in the series. If a series has 10 steps, then steps 1 .. 9 are input, and
the steps 2 .. 10 are truth. The network is trained to
minimize negative log probability of true observations given
predictions under variational approximation of product of
Gaussians.

\paragraph{Novelty detection} After training, the model is used to produce
predictions of future observations. There are two related but different
phenomena indication a novelty (anomaly) in time series behavior:
\begin{enumerate}
	\item Predicted volatility of the time series is high, that is, future
	observations can only be forecast uncertainly (with high variance). 
	\item Probability of actual observations, when observed, given a prediction
	from a past state, is low.
\end{enumerate}

Either phenomenon, or both of them, can be used to alert about novelties in the
time series. In recurrent neural network architectures, the hidden state ($h_t$
in Figure~\ref{fig:dt-time-series-models}) can be used to identify and classify
anomalies.

However, the basic scheme outlined above poses difficulty in applications with
high-dimensional time series and partially missing observations. Sampling based
uncertainty assertion impacts performance, and missing observations are often
imputed in \textit{ad hoc} manner. An architecture which
incorporates confidence about data and in which observed and
predicted data are interchangable is  highly, desirable. For example,k if out of 5 components 3 were
measured and 2 predicted from an earlier step we want to input
them both into the next time step for further forecasting. In
addition, the model architecture should be capable of robust uncertainty
prediction and benefit from training with multiple steps of out-of-sample data.
To achieve these objectives, we propose the following changes
(Figure~\ref{fig:dt-uncertainty-propagation}):

\begin{enumerate}
\item The input, as well as the output, is a vector of means followed
by standard deviations. If the data has 5 components, the input
will be 10-dimensional. For observed data --- measurements
present at the current time step --- the standard deviation is
zero. For missing data the input is mean and standard deviation
as predicted from the previous time step.

\item Since we introduce confidence into the input, we cannot train
the network myopically, on a single step prediction --- or the
standard deviations in the input data will always be zero, and
the network will never learn how to use them. To overcome this,
we train on multiple predicted steps. We feed each prediction,
without sampling, as input to the next step and compute the loss
as negative log probability of this number of future points versus our prediction.
\end{enumerate}

To illustrate, given the data set of 5 dimensions, the input has 10
dimensions. If we train with 3 time steps lookahead, the ground
truth will be a matrix of size 3x5. The prediction against
which the likelihood of this ground truth is computed will be
a matrix of size 3x10. Intuitively, we would expect the
predicted standard deviation to increase along the time axis
for each component.

The ability of Bayesian forecasting with uncertainty, in the form of a multivariate Gaussian distributions, far into the future, opens opportunity for application for more robust novelty detection
approaches. Instead of detecting novelty based on log probability of observations given predictions from the past, which is prone to false positives due to observation noise, novelties can be detected and analysed by comparing predictions of the same time point from different points in the past. In this case, KL-divergence between predictions provide a theoretically sound and robust mechanism for detection of anomalies~\cite{10.1145/3297280.3297414}, and is in particular relevant for monitoring of large OT environments with high dimensionality of time series and occasional missing values and heteroskedastic noise.

\section{Case Studies}

\section{Related Work}

\section{Discussion and Future Research}

Monitoring of complex systems based on multivariate time series meets
challenges in adapting a general algorithm to  domain-specific properties of
the particular system~\cite{DBLP:conf/iccst/DymshitsMT17}. As the \textbf{first
step}, the structure and scale of time series emerging in monitoring of power
plants, and relevance of various components of the time series for novelty
detection, in particular related to cybersecurity, has to be analyzed. As
compelling as it may sound, feeding all of the available observations in their
raw form to a general model will unlikely to result in robust novelty
detection. Hence, we propose to start with empirical evaluation of smoothing
and forecasting capabilities of existing multi-variate time series model on
different representations of the time series. Conventional neural multivariate
time series models, while having drawbacks, will aid in selecting the right
representation of the data, which can be assessed by the quality of
forecasting, which should be consistently high in normal operation and decrease
in presence of otherwise observed disturbances (hardware and operational
failures, cybersecurity attacks, etc.). Using established modelling approaches
as a test bench, we will be able to come with an appropriate representation of
the time series emerging in power infrastructure monitoring.

\textbf{Following this}, we propose implementing and evaluating the deep
Bayesian model, as outlined above, in its basic form, implemented as an
extension of an established time series prediction architecture. By doing so,
we will be able to use the earlier architecture as a comparison baseline, and
assess improvements due to Bayesian handling of uncertainty by comparing
performance of detection and identification of novelties. Both recorded real
and simulated novelties can be used for this stage.

Provided that the previous step succeeds, research on classification and
characterization of novelties can be commenced, which is, in a sense,
orthogonal to time series forecasting, and is based on extending the base model
with modules using the internal state of the recurrent network to learn to
recognize or characterize types and features of novelties. This layered plan of
the research allows to go back to earlier steps (data representation, extension
of time series models with Bayesian handling of uncertainty, advanced
techniques for novelty identification and characterization) in case results of
a later step uncover obstacles introduced earlier and require changes in theory
or implementation.

\end{document}

### 2022-02-15 paper.tex
intro in progress

@@
 \begin{abstract}
-Systems are commonly monitored for health and security through collection and
-streaming of multivariate time series. Advances in time series forecasting due
-to adoption of multilayer recurrent neural network architectures make it
-possible to forecast in high-dimensional time series, and identify and classify
-novelties early, based on subtle changes in the trends. However, mainstream
-approaches to multi-variate time series predictions do not handle well cases
-when the ongoing forecast must include uncertainty, nor they are robust to
-missing data. We introduce a new architecture for time series monitoring based
-on combination of state-of-the-art methods of forecasting in high-dimensional
-time series with full Bayesian handling of uncertainty. We demonstrate
-advantage of the architecture for time series forecasting and novelty
-detection, in particular with partially missing data, and empirically evaluate
-and compare the architecture to state-of-the-art approaches on a real-world
-data set.
+Systems are commonly monitored for health and security through
+collection and streaming of multivariate time series. Advances
+in time series forecasting due to adoption of multilayer
+recurrent neural network architectures make it possible to
+forecast in high-dimensional time series, and identify and
+classify novelties early, based on subtle changes in the trends.
+However, mainstream approaches to multi-variate time series
+predictions do not handle well cases when the ongoing forecast
+must include uncertainty, nor they are robust to missing data.
+We introduce a new architecture for time series monitoring based
+on combination of state-of-the-art methods of forecasting in
+high-dimensional time series with full Bayesian handling of
+uncertainty. We demonstrate advantage of the architecture for
+time series forecasting and novelty detection, in particular
+with partially missing data, and empirically evaluate and
+compare the architecture to state-of-the-art approaches on a
+real-world data set.
 \end{abstract}
@@
 
-Evolving state of OT environments  is monitored through collection and
-streaming of multivariate time series. Advances in time series forecasting due
-to adoption of multilayer recurrent neural network architectures make it
-possible to forecast in high-dimensional time series, and identify and
-classify novelties (anomalies) early, based on subtle changes in the trends.
-However, mainstream approaches to multi-variate time series predictions do not
-handle well cases when the ongoing forecast must include uncertainty and shape
-(possibly multimodal) of future distributions of observations. Stochastic
-latent variable variants of high-dimensional time series models where
-proposed, but so far have had to rely on sampling to account for uncertainty,
-limiting the performance of data handling. In addition, main-stream methods of
-multivariate time series forecasting are not well-suited for handling missing
-data, while transient unavailability of some of the data sources (e.g. when a
-sensor stops working, or a transport channel malfunctions) is  a common case
-with monitoring of complex systems.  Combining state-of-the-art methods of
-forecasting in high-dimensional time series (based on neural-network
-architectures) with full Bayesian handling of uncertainty (e.g. with help of
-deep probabilistic programming systems such as Pyro, Tensorflow Probability,
-and others) will allow to achieve robust forecasting (a precondition for
-anomaly detection) in high-dimensional time series with missing data.
-Representing future states as inferred posterior probability distributions
-will allow to identify and classify anomalies (e.g. hardware failures,
-software errors, communication, malicious activity, cyberattacks, e.t.c.) on a
-higher level of model abstraction — the latent variables of the model, and
-achieve finer and more robust anomaly detection and characterization.
+Modern information systems and operation environments are
+commonly monitored through collection and streaming of
+multivariate time series. The monitoring tasks comprise both
+forecasting, for planning of resource allocation and decision
+making, and novelty detection and characterization, for ensuring
+faultless functioning and early mitigation of failures and
+threats.  Advances in time series forecasting
+due to adoption of multilayer recurrent neural network
+architectures made it possible to forecast in high-dimensional
+time series, and identify and classify novelties (anomalies)
+early, based on subtle changes in the trends.  However,
+mainstream approaches to multi-variate time series modelling
+do not handle well cases when uncertainty is involved, either in
+the input, when some of the observations are missing, or in the
+output when the distribution of future observations, rather than
+their point values, is predicted. For forecast uncertainty
+modelling, stochastic latent variable
+variants of high-dimensional time series models where
+introduced, but so far have had to rely on sampling to account
+for uncertainty, limiting the performance of data handling.
+Imputation schemes were proposed for dealing with 
+missing data, however, they do not generally give a satisfactory
+solution in presence of transient unavailability of some of
+the data sources (e.g. when a sensor stops working, or a
+transport channel malfunctions), which is  a common case with
+monitoring of complex systems.
+
+A systematic and theoretically founded approach to handling
+both input and output uncertainty would thus constitute a
+significant and welcome contribution to the theory and practice
+of monitoring of multivariate time series. It would also be
+highly desirable for such approach to facilitate efficient
+offline (learning) and online (inference) computations. In this
+ongoing research, we propose a deep learning architecture which
+uses a simple but powerful extension of traditional recurrent
+neural network (RNN) architecture which allows both 
+\begin{itemize} 
+\item to handle missing inputs
+in some or all of the components in a multivariate time series,
+\item and to accomplish multi-step probabilistic forecasting
+\end{itemize}
+in high-dimensional time series, paving a path to better
+decision making and finer and more robust anomaly detection and
+characterization.  We evaluate the architecture on a real-world
+data set of multivariate time series collected from a 
+network for cloud computing, and empirically demonstrate
+advantage of the proposed architecture over commonly used
+approaches.
 

### 2022-02-15 paper.tex
in progress, next thing is refs

@@
 
-\title{Monitoring Time Series With Missing Values:\\a Deep Bayesian Approach}
+\title{Monitoring Time Series With Missing Values:\\a Deep Probabilistic Approach}
 
@@
 on combination of state-of-the-art methods of forecasting in
-high-dimensional time series with full Bayesian handling of
+high-dimensional time series with full probabilistic handling of
 uncertainty. We demonstrate advantage of the architecture for
@@
 
-\section{Concept}
+\section{Problem: Multivariate Time Series Forecasting}
+
+The core problem we address is forecasting in a multivariate
+time series. Formally, a \textit{time series} is a matrix $X$
+of shape $T\times N$, where $T$ is the number of time steps and
+$N$ is the number of dimensions. The time steps are assumed to
+be equispaced.  A $k$-step probabilistic \textit{forecast}
+$\mathcal{F}_{tk}$ at time $t$ is the belief distribution of
+time series $X_{t+1:t+k}$ for time steps $t+1 ... t+k$ given the
+observed time series $X_{1:t}$ for time steps $1...t$.
+
+The forecasting is accomplished by applying model $\mathcal{M}_\theta$
+parameterized by parameters $\theta$ to the observed time
+series:
+\begin{equation}
+\mathcal{F}_{tk} = \mathcal{M}_\theta(X_{1:t})
+\end{equation}
+The machine learning task is to devise $\theta^*$
+that gives the best forecast, in terms of a certain loss
+function. A natural loss in the probabilistic setting is the
+average negative log likelihood of $\theta$ given a training
+data set $\mathcal{X}$ of multiple time series:
+\begin{equation}
+	\theta^* = \arg\min_\theta \mathbb{E}_{X \in \mathcal{X},t
+	\in 1 ... T-k}\left[ -\log \Pr(X_{t+1:t+k}|M_\theta(X_{1:t}))\right]
+\end{equation}
+When the model is differentiable by $\theta$, the task is
+usually accomplished by performing a stochastic gradient loss
+minimization.
+
+In the basic case, $X$ is real-valued, $X \in \mathbb{R}^{T
+\times N}$. Here, we are interested in an extension of the basic
+case, in which some of the elements can be missing from $X$, that is
+$X \in (\mathbb{R} \cup \bot)^{T \times N}$.
+
+\section{Architecture: Recurrent Neural Network with Uncertainty Propagation}
 
@@
 
-There is a range of neural recurrent models of varying complexity to deal with
-time series prediction. Most models include a recurrent unit which threads the
-state through the time steps, accepts data as inputs and produces next step
-predictions as outputs. The simplest model is an RNN with a fully-connected
-readout layer to produce predictions (Figure~\ref{fig:dt-conventional-model}).
-RNN can be based on GRU, LSTM, or another variant, and is often multi-layer.
-Architectures  may also include intermediate modules and sampling-based
+A popular realization of the forecasting model
+$\mathcal{M}_\theta$ is a recurrent neural network (RNN), with
+$\theta$ corresponding to the network parameters.  There is a
+range of neural recurrent models of varying complexity to deal
+with time series forecasting. Most models include a recurrent
+unit which threads the state through the time steps, accepts
+data as inputs and produces next step predictions as outputs.
+The simplest model is an RNN with a fully-connected readout
+layer to produce forecasts
+(Figure~\ref{fig:dt-conventional-model}).  RNN can be based on
+LSTM~\cite{}, GRU~\cite{}, or another architectural variant, and
+is often multi-layer.  Architectures  may also include
+intermediate modules and sampling-based variational
+layers~\cite{}. The overall architecture stays almost the same,
+with more connections, intermediate modules and sampling-based
 variational layers.
 
-Variants are stochastic RNNs, variational RNNs, etc. The
-overall architecture stays almost the same, with more
-connections, intermediate modules and sampling-based variational
-layers.
+\paragraph{Input and Output} This architecture normally accepts
+data vectors and outputs vectors of predictions of means and
+standard deviations (the output is twice as wide as the input).
+Then, to generate a forecast, one samples from
+$$\mathcal{N}(\mu, \sigma)$$ for each component, and feeds the
+sample as the input to the next time step.
 
-\paragraph{Input and Output} This architecture normally accepts data vectors
-and outputs vectors of predictions of means and standard deviations (the output
-is twice as wide as the input). Then, to generate a forecast, one samples from
-$$\mathcal{N}(\mu, \sigma)$$ for each component, and feeds the sample as the
-input to the next time step.
+\paragraph{Training} The model is trained to maximize
+probability of prediction. In the most basic case, a single step
+is predicted for each time step in the series. If a series has
+10 steps, then steps 1 .. 9 are the input, and the steps 2 .. 10
+are the truth. The network is trained to minimize negative log
+probability of true observations given predictions under
+variational approximation of product of Gaussians.
 
-\paragraph{Training} The model is trained to maximize probability of
-prediction. In the most basic case, a single step is predicted for each time
-step in the series. If a series has 10 steps, then steps 1 .. 9 are input, and
-the steps 2 .. 10 are truth. The network is trained to
-minimize negative log probability of true observations given
-predictions under variational approximation of product of
-Gaussians.
-
-\paragraph{Novelty detection} After training, the model is used to produce
-predictions of future observations. There are two related but different
-phenomena indication a novelty (anomaly) in time series behavior:
+\paragraph{Novelty detection} After training, the model is used
+to produce predictions of future observations. There are two
+related but different phenomena indication a novelty (anomaly)
+in time series behavior:
 \begin{enumerate}
@@
 \end{enumerate}
+Either phenomenon, or both of them, can be used to alert about
+novelties in the time series. In recurrent neural network
+architectures, the hidden state ($h_t$ in
+Figure~\ref{fig:dt-time-series-models}) can be used to identify
+and classify anomalies.
 
-Either phenomenon, or both of them, can be used to alert about novelties in the
-time series. In recurrent neural network architectures, the hidden state ($h_t$
-in Figure~\ref{fig:dt-time-series-models}) can be used to identify and classify
-anomalies.
+However, the basic scheme outlined above poses difficulty in
+applications with high-dimensional time series and partially
+missing observations. Sampling based uncertainty assertion
+impacts performance, and missing observations are often imputed
+in \textit{ad hoc} manner~\cite{}. An architecture which
+incorporates confidence about data and in which observed and
+predicted data are interchangeable is  highly desirable. For
+example, if out of 5 components 3 were measured and 2 predicted
+from an earlier step we want to input them both into the next
+time step for further forecasting. In addition, the model
+architecture should be capable of robust uncertainty prediction
+and benefit from training with multiple steps of out-of-sample
+data.
 
-However, the basic scheme outlined above poses difficulty in applications with
-high-dimensional time series and partially missing observations. Sampling based
-uncertainty assertion impacts performance, and missing observations are often
-imputed in \textit{ad hoc} manner. An architecture which
-incorporates confidence about data and in which observed and
-predicted data are interchangable is  highly, desirable. For example,k if out of 5 components 3 were
-measured and 2 predicted from an earlier step we want to input
-them both into the next time step for further forecasting. In
-addition, the model architecture should be capable of robust uncertainty
-prediction and benefit from training with multiple steps of out-of-sample data.
-To achieve these objectives, we propose the following changes
-(Figure~\ref{fig:dt-uncertainty-propagation}):
+To achieve these objectives, \textbf{we propose the following
+changes} (Figure~\ref{fig:dt-uncertainty-propagation}):
 
@@
 present at the current time step --- the standard deviation is
-zero. For missing data the input is mean and standard deviation
-as predicted from the previous time step.
+zero. For missing data the input is the mean and the standard deviation
+as predicted from the preceding time steps.
 
-\item Since we introduce confidence into the input, we cannot train
-the network myopically, on a single step prediction --- or the
-standard deviations in the input data will always be zero, and
-the network will never learn how to use them. To overcome this,
-we train on multiple predicted steps. We feed each prediction,
-without sampling, as input to the next step and compute the loss
-as negative log probability of this number of future points versus our prediction.
+\item Since we introduce confidence into the input, we cannot
+train the network myopically, on a single step prediction --- or
+the standard deviations in the input data will always be zero,
+and the network will never learn how to use them. To overcome
+this, we train on multiple predicted steps. We feed each
+prediction, without sampling, as input to the next step and
+compute the loss as negative log probability of this number of
+future points versus our prediction.
 \end{enumerate}
@@
 dimensions. If we train with 3 time steps lookahead, the ground
-truth will be a matrix of size 3x5. The prediction against
+truth will be a matrix of size $3\times 5$. The prediction against
 which the likelihood of this ground truth is computed will be
-a matrix of size 3x10. Intuitively, we would expect the
+a matrix of size $3 \times 10$. Intuitively, we would expect the
 predicted standard deviation to increase along the time axis
@@
 
-The ability of Bayesian forecasting with uncertainty, in the form of a multivariate Gaussian distributions, far into the future, opens opportunity for application for more robust novelty detection
-approaches.
-Instead of detecting novelty based on log probability of observations given predictions from the past, which is prone to false positives due to observation noise, novelties can be detected and analysed by comparing predictions of the same time point from different points in the past.
-In this case, KL-divergence between predictions provide a theoretically sound and robust mechanism for detection of anomalies~\cite{10.1145/3297280.3297414}, and is in particular relevant for monitoring of large OT environments with high dimensionality of time series and occasional missing values and heteroskedastic noise.
+The ability of probabilistic forecasting with uncertainty, in
+the form of a multivariate Gaussian distributions, far into the
+future, opens opportunity for application for more robust
+novelty detection approaches. Instead of detecting novelty based
+on log probability of observations given predictions from the
+past, which is prone to false positives due to observation
+noise, novelties can be detected and analysed by comparing
+predictions of the same time point from different points in the
+past. In this case, KL-divergence between predictions provide a
+theoretically sound and robust mechanism for detection of
+anomalies, and is in particular relevant for monitoring of large
+operation environments with high dimensionality of time series
+and occasional missing values and heteroskedastic noise.
 
-\section{Case Studies}
+\section{Case Study: Monitoring a Computer Cloud}
 
@@
 
-Monitoring of complex systems based on multivariate time series meets
-challenges in adapting a general algorithm to  domain-specific properties of
-the particular system~\cite{DBLP:conf/iccst/DymshitsMT17}. As the \textbf{first
-step}, the structure and scale of time series emerging in monitoring of power
-plants, and relevance of various components of the time series for novelty
-detection, in particular related to cybersecurity, has to be analyzed. As
-compelling as it may sound, feeding all of the available observations in their
-raw form to a general model will unlikely to result in robust novelty
-detection. Hence, we propose to start with empirical evaluation of smoothing
-and forecasting capabilities of existing multi-variate time series model on
-different representations of the time series. Conventional neural multivariate
-time series models, while having drawbacks, will aid in selecting the right
-representation of the data, which can be assessed by the quality of
-forecasting, which should be consistently high in normal operation and decrease
-in presence of otherwise observed disturbances (hardware and operational
-failures, cybersecurity attacks, etc.). Using established modelling approaches
-as a test bench, we will be able to come with an appropriate representation of
-the time series emerging in power infrastructure monitoring.
+\section*{Acknowledgements}
 
-\textbf{Following this}, we propose implementing and evaluating the deep
-Bayesian model, as outlined above, in its basic form, implemented as an
-extension of an established time series prediction architecture. By doing so,
-we will be able to use the earlier architecture as a comparison baseline, and
-assess improvements due to Bayesian handling of uncertainty by comparing
-performance of detection and identification of novelties. Both recorded real
-and simulated novelties can be used for this stage.
-
-Provided that the previous step succeeds, research on classification and
-characterization of novelties can be commenced, which is, in a sense,
-orthogonal to time series forecasting, and is based on extending the base model
-with modules using the internal state of the recurrent network to learn to
-recognize or characterize types and features of novelties. This layered plan of
-the research allows to go back to earlier steps (data representation, extension
-of time series models with Bayesian handling of uncertainty, advanced
-techniques for novelty identification and characterization) in case results of
-a later step uncover obstacles introduced earlier and require changes in theory
-or implementation.
+\bibliographystyle{splncs04}
+\bibliography{refs.bib}
 

### 2022-02-15 paper.tex
in progress

@@
 	\in 1 ... T-k}\left[ -\log \Pr(X_{t+1:t+k}|M_\theta(X_{1:t}))\right]
+	\label{eqn:theta-star}
 \end{equation}
@@
 \paragraph{Input and Output} This architecture normally accepts
-data vectors and outputs vectors of predictions of means and
-standard deviations (the output is twice as wide as the input).
-Then, to generate a forecast, one samples from
-$$\mathcal{N}(\mu, \sigma)$$ for each component, and feeds the
-sample as the input to the next time step.
+observation vectors and outputs vectors of distribution
+parameters for the belief distribution of the observations at
+the next time step. In the case of frequently used normal
+distributions, the output vector consists of means $\mu$ and
+standard deviations $\sigma$ of all dimensions and is twice as
+wide as the input.  This allows to also assess uncertainty
+multiple steps into the future: one can repeatedly sample from
+the observation belief distribution at each future step, and
+feed the sample as the input to the following step.
 
@@
 is predicted for each time step in the series. If a series has
-10 steps, then steps 1 .. 9 are the input, and the steps 2 .. 10
-are the truth. The network is trained to minimize negative log
-probability of true observations given predictions under
-variational approximation of product of Gaussians.
+10 steps, then steps $1 ... 9$ are the input, and the steps $2
+... 10$ are the truth. Following~\eqref{eqn:theta-star}, the
+network is trained to minimize negative log probability of true
+observations given the predicted belief distributions.
 
@@
 \begin{enumerate}
-\item The input, as well as the output, is a vector of means followed
-by standard deviations. If the data has 5 components, the input
-will be 10-dimensional. For observed data --- measurements
-present at the current time step --- the standard deviation is
-zero. For missing data the input is the mean and the standard deviation
-as predicted from the preceding time steps.
+\item The input, as well as the output, is a vector of
+distribution parameters. For simplicity, we will confine further
+discussion to the case of independent normal belief
+distributions for each component; however, other distribution
+shapes can be used. For the independent normal
+distributions, the distribution parameter vector consists of the
+means followed by the standard deviations. If the data has 5
+components, the input will be 10-dimensional. For observed data
+--- measurements present at the current time step --- the
+standard deviation is zero. For missing data the input is the
+mean and the standard deviation as predicted from the preceding
+time steps.
 
@@
 The ability of probabilistic forecasting with uncertainty, in
-the form of a multivariate Gaussian distributions, far into the
-future, opens opportunity for application for more robust
+the form of multivariate normal distributions, far into the
+future, opens opportunity for application to more robust
 novelty detection approaches. Instead of detecting novelty based

### 2022-02-16 paper.tex
related work, acknowledgements

@@
 
+There appear to be two interconnected areas related to this
+research. One area is uncertainty representation and propagation
+in recurrent neural models. The other area is handling of
+missing values in time series, again in the context of
+recurrent neural models in particular.
+
+The importance of uncertainty quantification in deep learning is
+well understood~\cite{MFS+21}.  Recurrent neural networks can
+express forecast uncertainty through predicting distribution
+parameters, such as the mean and the standard deviation, instead
+of point values~\cite{HS97}. When expressing uncertainty by closed-form
+distributions is insufficient, stochastic latent
+variables are introduced into RNNs~\cite{YB21,CKD+15,FSK+16}. 
+Uncertainty representation in RNNs is related to uncertainty
+propagation and multi-step forecasting. For multi-step
+forecasting, uncertainty must be propagated multiple steps into
+the future. Uncertainty propagation is usually achieved through
+random sampling during training or inference~\cite{LYY+19,AV20,YB21}. 
+Our approach differs in that conventional RNN architectures are
+leveraged to represent uncertainty in both the input and
+the output, and that uncertainty propagation is accomplished
+deterministically, without resorting to random sampling, which
+facilitates efficient training and inference.
+
+Handling of missing values in time series has inspired research
+for decades due to the fact that many otherwise efficient and
+robust algorithms, in particular those based on recurrent neural
+architectures, require that all values in the time series
+are present and lie within a valid range~\cite{WSY+21}. A
+widespread approach is to \textit{impute} the data,
+that is, to replace missing values with values inferred from
+other values in the same time series or in other time series
+in the data set~\cite{KC18,SYG+19}. Alternatively, a
+missing value is treated as an observation itself, often by
+introducing an auxiliary indicator variable~\cite{LKW16,BDS21}.
+In our work, we take a third approach --- a missing value,
+either due to an absent observation or in the course of
+multi-step forecasting, is replaced by a parametrically
+specified belief distribution of the value based on the past
+observations.
+
 \section{Discussion and Future Research}
@@
 \section*{Acknowledgements}
+
+We thank PUB+ for providing computational facilities for
+conducting the empirical evaluation. David Tolpin is partially
+supported by Israel-U.S. Industrial Research and Development
+Foundation's Cybersecurity technology for critical power
+infrastructure AI-based centralized defense and edge resilience
+project.
+
+\nocite{BDS21}
+\nocite{CPK+18}
+\nocite{SYG+19}
+\nocite{KC18}
+\nocite{LKW16}
+\nocite{BRH15}
+\nocite{AV20}
+\nocite{FSK+16}
+\nocite{MFS+21}
+\nocite{FSP+16}
+\nocite{SZN+19}
+\nocite{LYY+19}
+\nocite{CYP+21}
+\nocite{YB21}
+\nocite{CKD+15}
+\nocite{HS97}
+\nocite{CMB+14}
+\nocite{WSY+21}
 

### 2022-02-16 paper.tex
minor

@@
 supported by Israel-U.S. Industrial Research and Development
-Foundation's Cybersecurity technology for critical power
-infrastructure AI-based centralized defense and edge resilience
+Foundation's \textit{Cybersecurity technology for critical power
+infrastructure AI-based centralized defense and edge resilience}
 project.

### 2022-02-17 paper.tex
discussion and future

@@
 
+We presented a deep probabilistic architecture for uncertainty
+propagation in multivariate time series. This architecture
+organically handles two important problems in deep time series
+modelling: missing data and multi-step forecasting. Empirical
+evaluation demonstrated that our approach outperforms
+conventional baselines in terms of forecasting accuracy, while
+still being easy to implement. Since, unlike some other
+approaches to uncertainty propagation, our architecture
+avoids sampling, uncertainty can be propagated efficiently and
+represented in closed parametric form, rather than approximated
+by samples and posterior intervals.
+
+We confined most of the discussion to the normal uncertainty shape.
+Other distributions can be used instead of the normal
+distributions where appropriate, provided their parameterization
+allows to express a certain observation as well as an uncertain
+belief. Analysis of distributions for representing uncertainty
+and their feasible parameterization is a subject of ongoing
+research. Another research direction worth exploring is
+extension of the presented architecture to bidirectional
+recurrent neural networks~\cite{BRH15}. Bidirectional RNNs allow
+to account for both past and future observations where
+appropriate, but make apparently make uncertainty propagation
+more complicated. Still, preliminary results suggest that
+uncertainty in bidirectional RNNs can be handled in a similar
+manner, further facilitating efficient probabilistic uncertainty
+propagation in a broader class of deep learning models for time
+series.
+
 \section*{Acknowledgements}

### 2022-02-17 paper.tex
first submittable version

@@
 
+We evaluate the proposed architecture on a data set of
+monitoring a cluster of 100 computing nodes in the cloud. For
+each node, the incoming and the outgoing network traffic (in
+bytes) and the CPU usage (relative) are logged with 1 minute
+resolution. 240 hours were logged, resulting in 12000 120-minute
+3-dimensional samples.  We split the dataset into the training,
+validation, and test as 80\%, 10\%, and 10\% correspondingly.
+Since the original data set does not have any missing data
+points, we emulated data sets with missing data by randomly
+removing 5\%, 10\%, 20\%, and 50\% of the data.
+
+We used a 3-layer GRU-based recurrent neural network with hidden
+size 64 and 20\% dropout between layers. We trained the network
+with lookahead depths (number of steps to forecast in the
+future) 2, 4, 8, and 16 using the Adam optimizer with learning
+rate 0.001, training for 20 epochs (sufficient for convergence).
+We performed the training on a cloud computing node with 1
+NVIDIA T4 GPU, 4 Intel Xeon Platinum CPUs, and 64 Gb memory. The
+training of a single model took 20 minutes.
+
+We compared our approach with conventional imputation methods
+`replace by the mean' and `replace by a random sample'.
+In the `replace by mean' method, a missing value is replaced 
+by the mean of the forecast. In the `replace by a random
+sample' method, a missing value is replaced by a random sample
+drawn from the forecast. As a performance metrics, we
+used per-point negative log-likelihood loss on the test set. 
+Tables~\ref{tab:replace-by-mean} and~\ref{tab:replace-by-sample}
+show the difference in loss between uncertainty propagation and
+`replace by the mean' and `replace by a random sample',
+correspondingly. The greater is the number, the worse is the
+forecasting by each of the methods compared to uncertainty
+propagation.
+\begin{table}[t]
+\caption{Uncertainty propagation vs. `replace by the mean`.}
+\label{tab:replace-by-mean}
+	\setlength\tabcolsep{18pt}
+	\centering
+	\begin{tabular}{r|c c c c}
+	 \textbf{missing} & \textbf{2} & \textbf{4} & \textbf{8} & \textbf{16} \\ \hline
+		  \textbf{5\%} & 0.001 & 0.001 &  0.06 & 0.10  \\
+		 \textbf{10\%} & 0.001 & 0.002 &  0.08 & 0.11  \\
+		 \textbf{20\%} & 0.003 & 0.003 &  0.11 & 0.13  \\
+		 \textbf{50\%} & 0.004 & 0.006 &  0.12 & 0.16  
+	\end{tabular}
+\end{table}
+\begin{table}[t]
+\caption{Uncertainty propagation vs. `replace by a random sample`.}
+\vspace{1em}
+\label{tab:replace-by-sample}
+	\setlength\tabcolsep{18pt}
+	\centering
+	\begin{tabular}{r|c c c c}
+	 \textbf{missing} & \textbf{2} & \textbf{4} & \textbf{8} & \textbf{16} \\ \hline
+		  \textbf{5\%} & 0.04 & 0.04 & 0.13 & 0.14 \\
+		 \textbf{10\%} & 0.06 & 0.06 & 0.16 & 0.24 \\
+		 \textbf{20\%} & 0.11 & 0.12 & 0.20 & 0.27 \\
+		 \textbf{50\%} & 0.18 & 0.19 & 0.28 & 0.30   
+	\end{tabular}
+\end{table}
+One can see that in all cases uncertainty propagation provides
+better forecasts than either of the conventional methods.
+
+The code and data for the case studies are available at
+\url{https://bitbucket.org/dtolpin/dbts-studies/}.
+
 \section{Related Work}

### 2022-02-17 paper.tex
added example

@@
 \usepackage[T1]{fontenc}
+\usepackage{graphicx}
 \usepackage{float}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{subfigure}{.64\textwidth}
@@
 
-We used a 3-layer GRU-based recurrent neural network with hidden
-size 64 and 20\% dropout between layers. We trained the network
-with lookahead depths (number of steps to forecast in the
-future) 2, 4, 8, and 16 using the Adam optimizer with learning
-rate 0.001, training for 20 epochs (sufficient for convergence).
-We performed the training on a cloud computing node with 1
-NVIDIA T4 GPU, 4 Intel Xeon Platinum CPUs, and 64 Gb memory. The
-training of a single model took 20 minutes.
-
-We compared our approach with conventional imputation methods
-`replace by the mean' and `replace by a random sample'.
-In the `replace by mean' method, a missing value is replaced 
-by the mean of the forecast. In the `replace by a random
-sample' method, a missing value is replaced by a random sample
-drawn from the forecast. As a performance metrics, we
-used per-point negative log-likelihood loss on the test set. 
-Tables~\ref{tab:replace-by-mean} and~\ref{tab:replace-by-sample}
-show the difference in loss between uncertainty propagation and
-`replace by the mean' and `replace by a random sample',
-correspondingly. The greater is the number, the worse is the
-forecasting by each of the methods compared to uncertainty
-propagation.
 \begin{table}[t]
@@
 \end{table}
+We used a 3-layer GRU-based recurrent neural network with hidden
+size 64 and 20\% dropout between layers. We trained the network
+with lookahead depths (number of steps to forecast in the
+future) 2, 4, 8, and 16 using the Adam optimizer with learning
+rate 0.001, training for 20 epochs (sufficient for convergence).
+We performed the training on a cloud computing node with 1
+NVIDIA T4 GPU, 4 Intel Xeon Platinum CPUs, and 64 Gb memory. The
+training of a single model took 20 minutes.
+
+We compared our approach with conventional imputation methods
+`replace by the mean' and `replace by a random sample'.
+In the `replace by mean' method, a missing value is replaced 
+by the mean of the forecast. In the `replace by a random
+sample' method, a missing value is replaced by a random sample
+drawn from the forecast. As a performance metrics, we
+used per-point negative log-likelihood loss on the test set. 
+Tables~\ref{tab:replace-by-mean} and~\ref{tab:replace-by-sample}
+show the difference in loss between uncertainty propagation and
+`replace by the mean' and `replace by a random sample',
+correspondingly. The greater is the number, the worse is the
+forecasting by each of the methods compared to uncertainty
+propagation.
 One can see that in all cases uncertainty propagation provides
 better forecasts than either of the conventional methods.
+
+\begin{figure}[t]
+\includegraphics[width=0.9\linewidth]{dist-vs-mean.pdf}
+\caption{Uncertainty propagation vs `replace by mean'. 95\% confidence intervals are shaded.}
+\label{fig:dist-vs-mean}
+\end{figure}
+As an illustration of the advantage of uncertainty propagation,
+consider Figure~\ref{fig:dist-vs-mean}, which shows forecasts
+using uncertainty propagation and `replace by mean' in presence
+of missing values. Forecasts through uncertainty propagation
+result in adequate confidence intervals. However, when missing
+values are replaced by the mean of the belief distribution,
+further forecasts are overconfident and too many observations
+fall outside of 95\% confidence intervals.
 

### 2022-02-17 paper.tex
cosmetic

@@
 (Figure~\ref{fig:dt-conventional-model}).  RNN can be based on
-LSTM~\cite{}, GRU~\cite{}, or another architectural variant, and
+LSTM~\cite{HS97}, GRU~\cite{CMB+14}, or another architectural variant, and
 is often multi-layer.  Architectures  may also include
 intermediate modules and sampling-based variational
-layers~\cite{}. The overall architecture stays almost the same,
-with more connections, intermediate modules and sampling-based
-variational layers.
+layers~\cite{CKD+15,YB21}. The overall architecture stays almost
+the same, with more connections, intermediate modules and
+sampling-based variational layers.
 
@@
 impacts performance, and missing observations are often imputed
-in \textit{ad hoc} manner~\cite{}. An architecture which
+heuristically~\cite{LKW16,SYG+19}. An architecture which
 incorporates confidence about data and in which observed and

### 2022-02-18 paper.tex
for submission

@@
 However, mainstream approaches to multi-variate time series
-predictions do not handle well cases when the ongoing forecast
+predictions do not handle well cases when the ongoing forecasts
 must include uncertainty, nor they are robust to missing data.
@@
 
-\begin{figure}[t]
+We introduce here a recurrent neural network architecture which
+facilitates uncertainty propagation. The architecture is capable
+both of handling missing values and of multi-step forecasting.
+We begin with description of conventional forecasting with RNNs.
+Then, we describe our proposed architecture as an extension to
+the conventional model.
+
+\begin{figure}
 \begin{subfigure}{.64\textwidth}
@@
 
+\subsection{Conventional Forecasting}
+
 A popular realization of the forecasting model
@@
 is often multi-layer.  Architectures  may also include
-intermediate modules and sampling-based variational
+intermediate modules, and sampling-based variational
 layers~\cite{CKD+15,YB21}. The overall architecture stays almost
@@
 parameters for the belief distribution of the observations at
-the next time step. In the case of frequently used normal
-distributions, the output vector consists of means $\mu$ and
-standard deviations $\sigma$ of all dimensions and is twice as
-wide as the input.  This allows to also assess uncertainty
-multiple steps into the future: one can repeatedly sample from
-the observation belief distribution at each future step, and
-feed the sample as the input to the following step.
+the next time step. In the simplest case, the network produces a 
+single output for each input, that is the dimensions of the
+input and the output vector coincide. This corresponds to the
+assumption of homoskedasticity of epistemic noise, and either
+the mean squared error (corresponding to the Gaussian error
+distribution) or the mean absolute error (corresponding to the
+Laplace error distribution) is minimized.
+
+More generally though, the epistemic noise is better modelled
+heteroskedastically, and a two-parameter loss distribution,
+with the location and the scale as the parameters.
+In the case of the frequently used normal (Gaussian)
+distribution, the output vector consists of means $\mu$
+(location) and standard deviations $\sigma$ (scale) of all
+dimensions and is twice as wide as the input.
 
 \paragraph{Training} The model is trained to maximize
-probability of prediction. In the most basic case, a single step
-is predicted for each time step in the series. If a series has
-10 steps, then steps $1 ... 9$ are the input, and the steps $2
-... 10$ are the truth. Following~\eqref{eqn:theta-star}, the
+probability of prediction. In the most basic case, called
+\textit{out-of-sample one-step} forecasting, a single step
+is predicted for each time step in the series. In an $n$-step
+time series, steps $1 ... n-1$ are used as the input, and steps
+$1 ... n$ as the ground truth. Following~\eqref{eqn:theta-star}, the
 network is trained to minimize negative log probability of true
-observations given the predicted belief distributions.
+observations given the predicted belief distributions. More
+generally, a model can also be trained to predict more than a
+single step at once into the future, however this is rarely used in
+practice because the necessary size of the training data set
+grows exponentially with the prediction depth. Instead, future
+predictions are produced recurrently during forecasting.
 
-\paragraph{Novelty detection} After training, the model is used
-to produce predictions of future observations. There are two
-related but different phenomena indication a novelty (anomaly)
+\paragraph{Forecasting} Forecasting is accomplished by passing
+past observations through the model to obtain forecasts for the
+future time steps. In the \textit{out-of-sample one-step} mode,
+a single step into the future is forecast. If a longer forecast
+is required, the current forecast is entered as the input at the
+next time step, time after time, up to the required length.
+Either the location (the point forecast) or a random sample from
+the belief distribution is used as the future input. Using
+random samples also allows to assess uncertainty multiple steps
+into the future: one can repeatedly sample from the belief
+distribution at each future step, and feed the sample as the
+input to the following step. Then, based on produced samples at
+future steps, one can estimate uncertainty intervals. Such
+Monte-Carlo handling of uncertainty is quite expensive
+computationally though, because the standard deviation of
+prediction error decreases as slowly as $\sqrt{N}$
+with the number of samples $N$, on one hand, and uncertainty
+may, in general, grow exponentially with prediction depth, on
+the other hand.
+
+\paragraph{Novelty detection} Forecasts produced by the model
+can be used for a number of purposes, including decision making
+and, in particular, novelty (anomaly) detection. There are two
+related but different phenomena indicating a novelty
 in time series behavior:
@@
 
-However, the basic scheme outlined above poses difficulty in
+\subsection{Forecasting with Uncertainty Propagation}
+
+The basic scheme outlined above poses difficulty in
 applications with high-dimensional time series and partially
@@
 example, if out of 5 components 3 were measured and 2 predicted
-from an earlier step we want to input them both into the next
+from an earlier step we want to input all of them into the next
 time step for further forecasting. In addition, the model
@@
 
-To achieve these objectives, \textbf{we propose the following
-changes} (Figure~\ref{fig:dt-uncertainty-propagation}):
+Our proposed architecture is based on the observation that if
+(at least) the location and the scale are used  to represent
+forecasts, an observation (that is, certain knowledge at a given
+step) can also be expressed using two parameters, by setting the
+location to the observation, and the scale to 0. For the normal
+distribution $\mathcal{N}(\mu, \sigma)$, the location and scale
+parameterization is straightforward, corresponding to $\mu$ and
+$\sigma$, however other belief distributions can be parameterized
+by location and scale as easily, e.g. the log-normal,
+Gamma, or Laplace distribution. For conciseness, we will confine
+further discussion to the case of independent normal belief
+distributions for each component; however, other distribution
+shapes can also be used. Based on this observation,
+\textbf{we propose the following
+extension} to the conventional RNN-based forecasting model
+(Figure~\ref{fig:dt-uncertainty-propagation}):
 
 \begin{enumerate}
-\item The input, as well as the output, is a vector of
-distribution parameters. For simplicity, we will confine further
-discussion to the case of independent normal belief
-distributions for each component; however, other distribution
-shapes can be used. For the independent normal
+\item \textit{The input, as well as the output, is a vector of
+distribution parameters}. For the independent normal
 distributions, the distribution parameter vector consists of the
@@
 
-\item Since we introduce confidence into the input, we cannot
-train the network myopically, on a single step prediction --- or
+\item  Training can, in principle, be accomplished on data with
+missing values, but training on data with missing values
+incurs performance drawbacks and should be avoided. First,
+handling missing values and replacing them with early
+predictions introduces contingency in the forward run of the RNN
+and slows down significantly the execution during training.
+Second, missing values should, in general, themselves be viewed
+as anomalies. One must be able to handle them during inference, 
+but should not rely on their presence in the training data.
+
+Therefore, we devise a scheme for \textit{training our model on data
+that does not contain missing values}. Even in applications where
+missing values are common in inference, training data without
+missing values is usually readily available. However,
+since we introduce confidence into the input, we cannot
+train the network myopically, in out-of-sample one-step manner --- 
 the standard deviations in the input data will always be zero,
@@
 the form of multivariate normal distributions, far into the
-future, opens opportunity for application to more robust
-novelty detection approaches. Instead of detecting novelty based
-on log probability of observations given predictions from the
-past, which is prone to false positives due to observation
-noise, novelties can be detected and analysed by comparing
-predictions of the same time point from different points in the
-past. In this case, KL-divergence between predictions provide a
-theoretically sound and robust mechanism for detection of
-anomalies, and is in particular relevant for monitoring of large
-operation environments with high dimensionality of time series
-and occasional missing values and heteroskedastic noise.
+future, opens opportunity for application to more robust novelty
+detection approaches. Instead of detecting novelty based on log
+probability of observations given predictions from the
+past~\cite{CBK09}, which is prone to false positives due to
+observation noise, novelties can be detected and analysed by
+comparing predictions of the same time point from different
+points in the past.  In this case,
+KL-divergence between predictions provides a theoretically sound
+and robust mechanism for detection of anomalies, and is in
+particular relevant for monitoring of large operation
+environments with high dimensionality of time series and
+occasional missing values and heteroskedastic noise~\cite{ASH08,T19}.
 
@@
 validation, and test as 80\%, 10\%, and 10\% correspondingly.
-Since the original data set does not have any missing data
+Since the original data set does not have many missing data
 points, we emulated data sets with missing data by randomly
@@
 
-\begin{table}[t]
+We used a 3-layer GRU-based recurrent neural network with hidden
+size 64 and 20\% dropout between layers. We trained the network
+with lookahead depths (number of steps to forecast in the
+future) 2, 4, 8, and 16 using the Adam optimizer with learning
+rate 0.001, training for 20 epochs (sufficient for convergence).
+We performed the training on a cloud computing node with 1
+NVIDIA T4 GPU, 4 Intel Xeon Platinum CPUs, and 64 Gb memory. The
+training of a single model took 20 minutes.
+
+\begin{table}
 \caption{Uncertainty propagation vs. `replace by the mean`.}
@@
 \end{table}
-\begin{table}[t]
+\begin{table}
 \caption{Uncertainty propagation vs. `replace by a random sample`.}
@@
 \end{table}
-We used a 3-layer GRU-based recurrent neural network with hidden
-size 64 and 20\% dropout between layers. We trained the network
-with lookahead depths (number of steps to forecast in the
-future) 2, 4, 8, and 16 using the Adam optimizer with learning
-rate 0.001, training for 20 epochs (sufficient for convergence).
-We performed the training on a cloud computing node with 1
-NVIDIA T4 GPU, 4 Intel Xeon Platinum CPUs, and 64 Gb memory. The
-training of a single model took 20 minutes.
-
 We compared our approach with conventional imputation methods
 `replace by the mean' and `replace by a random sample'.
-In the `replace by mean' method, a missing value is replaced 
+In the `replace by the mean' method, a missing value is replaced 
 by the mean of the forecast. In the `replace by a random
@@
 
-\begin{figure}[t]
-\includegraphics[width=0.9\linewidth]{dist-vs-mean.pdf}
-\caption{Uncertainty propagation vs `replace by mean'. 95\% confidence intervals are shaded.}
+\begin{figure}
+\centering
+\includegraphics[width=0.95\linewidth]{dist-vs-mean.pdf}
+\caption{Uncertainty propagation vs `replace by the mean'. 95\% confidence intervals are shaded.}
 \label{fig:dist-vs-mean}
@@
 consider Figure~\ref{fig:dist-vs-mean}, which shows forecasts
-using uncertainty propagation and `replace by mean' in presence
+using uncertainty propagation and `replace by the mean' in presence
 of missing values. Forecasts through uncertainty propagation
@@
 to account for both past and future observations where
-appropriate, but make apparently make uncertainty propagation
+appropriate, but apparently make uncertainty propagation
 more complicated. Still, preliminary results suggest that

### 2022-03-15 paper.tex

@@
 More generally though, the epistemic noise is better modelled
-heteroskedastically, and a two-parameter loss distribution,
+heteroskedastically, using a two-parameter loss distribution,
 with the location and the scale as the parameters.
